---
title: "Academic Figures Quiz"
subtitle: "Test Your Knowledge of Figure Best Practices"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(shiny)
knitr::opts_chunk$set(echo = FALSE)

# Initialize a reactive value to track score
# Note: learnr provides built-in progress tracking
```

## Introduction

Welcome to the Academic Figures Quiz! This quiz tests your knowledge based on the "Academic Figures: Common Pitfalls and Best Practices" presentation.

**Topics covered:**

- Color gradients and palettes (7 questions)
- Heatmap scaling (3 questions)
- File formats (5 questions)
- Text sizing (2 questions)
- Themes and styling (2 questions)
- Saving plots (3 questions)
- Post-processing (2 questions)
- Factor ordering (3 questions)
- Interactive plots (4 questions)

**Total: 31 questions**

Good luck! 🎯

---

## Section 1: Color Gradients

### Question 1

```{r q1}
quiz(
  question("What is the main problem with rainbow color scales?",
    answer("They are too colorful"),
    answer("They are perceptually non-uniform", correct = TRUE, message = "Correct! Rainbow scales create false boundaries because colors are not perceived uniformly."),
    answer("They use too many colors"),
    answer("They are not available in R"),
    allow_retry = TRUE
  )
)
```

### Question 2

```{r q2}
quiz(
  question("Which color scale is recommended as the DEFAULT for continuous data?",
    answer("Rainbow"),
    answer("Jet"),
    answer("Viridis", correct = TRUE, message = "Correct! Viridis is perceptually uniform, colorblind-safe, and prints well in grayscale."),
    answer("Heat colors"),
    allow_retry = TRUE
  )
)
```

### Question 3

```{r q3}
quiz(
  question("What percentage of men have some form of color blindness (deuteranopia)?",
    answer("~1%"),
    answer("~4%"),
    answer("~8%", correct = TRUE, message = "Correct! About 8% of men have deuteranopia (green-blind vision)."),
    answer("~15%"),
    allow_retry = TRUE
  )
)
```

### Question 4

```{r q4}
quiz(
  question("Which of these color scales remain readable when desaturated (grayscale)?",
    answer("Rainbow"),
    answer("Jet"),
    answer("Viridis", correct = TRUE),
    answer("Magma", correct = TRUE),
    answer("Cividis", correct = TRUE),
    correct = "Excellent! Viridis family scales are designed to maintain information in grayscale.",
    allow_retry = TRUE
  )
)
```

### Question 5

```{r q5}
quiz(
  question("What medical consequence was found in Borkin et al.'s study on heart disease diagnosis?",
    answer("Physicians using jet colormap made more errors and were slower", correct = TRUE, message = "Correct! The jet colormap caused physicians to misinterpret medical images, leading to diagnostic errors."),
    answer("Rainbow scales improved diagnosis speed"),
    answer("Color scales had no effect on diagnosis"),
    answer("All color scales performed equally"),
    allow_retry = TRUE
  )
)
```

### Question 6

```{r q6}
quiz(
  question("Which viridis option is specifically optimized for colorblind viewers?",
    answer("viridis"),
    answer("magma"),
    answer("plasma"),
    answer("cividis", correct = TRUE, message = "Correct! Cividis was designed to be equally accessible to people with and without color vision deficiency."),
    allow_retry = TRUE
  )
)
```

### Question 7

```{r q7}
quiz(
  question("When should you use diverging color scales?",
    answer("For all continuous data"),
    answer("When data has a meaningful center point (like zero)", correct = TRUE, message = "Correct! Diverging scales (like RdBu) are best for data with a meaningful midpoint, such as fold changes or correlations."),
    answer("Only for categorical data"),
    answer("Never use diverging scales"),
    allow_retry = TRUE
  )
)
```

---

## Section 2: Heatmaps

### Question 8

```{r q8}
quiz(
  question("What happens when you have outliers in a heatmap?",
    answer("The outlier compresses the color scale for all other values", correct = TRUE, message = "Correct! Extreme outliers dominate the color scale, making it hard to see patterns in the rest of the data."),
    answer("Outliers have no effect on visualization"),
    answer("Outliers improve contrast"),
    answer("R automatically removes outliers"),
    allow_retry = TRUE
  )
)
```

### Question 9

```{r q9}
quiz(
  question("Which scaling method is most robust to outliers for heatmaps?",
    answer("Z-score (mean ± SD)"),
    answer("MAD scaling (median ± MAD)", correct = TRUE, message = "Correct! MAD (Median Absolute Deviation) scaling is robust to outliers since it uses median instead of mean."),
    answer("No scaling"),
    answer("Log transformation only"),
    allow_retry = TRUE
  )
)
```

### Question 10

```{r q10}
quiz(
  question("What is quantile capping in heatmaps?",
    answer("Removing all data above the median"),
    answer("Setting extreme values to specific percentile thresholds (e.g., 5th and 95th)", correct = TRUE, message = "Correct! Quantile capping limits extreme values to specific percentiles while preserving the overall distribution."),
    answer("Deleting outlier samples entirely"),
    answer("Converting all values to quantiles"),
    allow_retry = TRUE
  )
)
```

---

## Section 3: File Formats

### Question 11

```{r q11}
quiz(
  question("Which format should you use for plots (lines, text, shapes)?",
    answer("JPEG"),
    answer("PNG"),
    answer("PDF or SVG", correct = TRUE, message = "Correct! Vector formats (PDF/SVG) are perfect for plots because they scale infinitely without quality loss."),
    answer("BMP"),
    allow_retry = TRUE
  )
)
```

### Question 12

```{r q12}
quiz(
  question("Which format should you use for actual photographs (camera/microscope images)?",
    answer("PDF"),
    answer("SVG"),
    answer("PNG for web, TIFF for publication", correct = TRUE, message = "Correct! TIFF (300+ DPI) for publication quality, PNG/JPEG/WebP (150+ DPI) for presentations and web. Avoid JPEG for scientific images due to compression artifacts."),
    answer("EPS"),
    allow_retry = TRUE
  )
)
```

### Question 13

```{r q13}
quiz(
  question("What is the main disadvantage of JPEG for scientific figures?",
    answer("It's too large"),
    answer("It uses lossy compression that creates artifacts", correct = TRUE, message = "Correct! JPEG's lossy compression can create visible artifacts around text and sharp edges."),
    answer("It doesn't support color"),
    answer("It can't be opened in most programs"),
    allow_retry = TRUE
  )
)
```

### Question 14

```{r q14}
quiz(
  question("You have a plot with a smooth color gradient generated in R. What format should you use?",
    answer("PDF or SVG - keep it as vector if possible", correct = TRUE, message = "Correct! If you created it in code (R/Python), save as vector (PDF/SVG) for infinite scalability. Only use raster (TIFF) if the journal doesn't accept vector formats."),
    answer("Always JPEG"),
    answer("Always PNG"),
    answer("BMP"),
    allow_retry = TRUE
  )
)
```

### Question 15

```{r q15}
quiz(
  question("What DPI (dots per inch) is recommended for high-quality print figures?",
    answer("72 dpi"),
    answer("150 dpi"),
    answer("300 dpi or higher", correct = TRUE, message = "Correct! 300 dpi is the standard for print quality. 600 dpi is even better for critical publications."),
    answer("1000 dpi"),
    allow_retry = TRUE
  )
)
```

---

## Section 4: Text and Element Sizing

### Question 16

```{r q16}
quiz(
  question("What is the recommended way to make text appear larger in a plot?",
    answer("Increase font size for every element manually"),
    answer("Use smaller figure dimensions so text appears larger relative to plot", correct = TRUE, message = "Correct! Reducing canvas size makes text appear proportionally larger without manual adjustments."),
    answer("Always use 24pt font"),
    answer("Zoom in on the plot"),
    allow_retry = TRUE
  )
)
```

### Question 17

```{r q17}
quiz(
  question("What does the base_size parameter control in ggplot2 themes?",
    answer("The size of the entire plot"),
    answer("The base font size (all text scales relative to this)", correct = TRUE, message = "Correct! base_size controls the base font size, and all other text elements scale relative to it."),
    answer("Only the title size"),
    answer("The point size in scatter plots"),
    allow_retry = TRUE
  )
)
```

---

## Section 5: Themes and Styling

### Question 18

```{r q18}
quiz(
  question("How do you set a default theme for all plots in your R session?",
    answer("ggplot2::set_theme()"),
    answer("theme_set(theme_bw())", correct = TRUE, message = "Correct! theme_set() applies a theme to all subsequent plots in the session."),
    answer("options(ggplot.theme = theme_bw())"),
    answer("You can't set a default theme"),
    allow_retry = TRUE
  )
)
```

### Question 19

```{r q19}
quiz(
  question("What is the advantage of using the ggpubr package?",
    answer("It provides publication-ready themes and statistical comparisons", correct = TRUE, message = "Correct! ggpubr offers clean themes and easy-to-add statistical comparisons for publication."),
    answer("It makes plots render faster"),
    answer("It's required for all ggplot2 plots"),
    answer("It only works with base R plots"),
    allow_retry = TRUE
  )
)
```

---

## Section 6: Saving Plots

### Question 20

```{r q20}
quiz(
  question("What is the recommended function for saving ggplot2 plots?",
    answer("pdf() ... dev.off()"),
    answer("ggsave()", correct = TRUE, message = "Correct! ggsave() is the modern, user-friendly way to save ggplot2 plots."),
    answer("save()"),
    answer("write.plot()"),
    allow_retry = TRUE
  )
)
```

### Question 21

```{r q21}
quiz(
  question("Why should you specify the plot object explicitly in ggsave()?",
    answer("It's required by R"),
    answer("To ensure you save the intended plot, not the last one displayed", correct = TRUE, message = "Correct! Explicitly specifying the plot prevents accidentally saving the wrong plot."),
    answer("It makes the file smaller"),
    answer("It's faster"),
    allow_retry = TRUE
  )
)
```

### Question 22

```{r q22}
quiz(
  question("What is the advantage of using Cairo devices (device = 'cairo_pdf')?",
    answer("Files are smaller"),
    answer("Better font embedding and Unicode support", correct = TRUE, message = "Correct! Cairo devices provide better font handling and support for special characters."),
    answer("Plots render faster"),
    answer("Required for all PDFs"),
    allow_retry = TRUE
  )
)
```

---

## Section 7: Post-Processing

### Question 23

```{r q23}
quiz(
  question("What is the recommended approach for making small edits to plots?",
    answer("Edit in PowerPoint and save"),
    answer("Make changes in R code and re-export", correct = TRUE, message = "Correct! Always maintain R scripts as the source of truth for reproducibility."),
    answer("Edit in Photoshop"),
    answer("Edit the PDF directly"),
    allow_retry = TRUE
  )
)
```

### Question 24

```{r q24}
quiz(
  question("What is the problem with 'Copy as Metafile' from RStudio?",
    answer("It's not a vector format"),
    answer("It corrupts plot symbols", correct = TRUE, message = "Correct! Copy as Metafile often distorts plot symbols (e.g., round points become crosses with rings)."),
    answer("It only works on Windows"),
    answer("It compresses the image"),
    allow_retry = TRUE
  )
)
```

---

## Section 8: Factor Ordering

### Question 25

```{r q25}
quiz(
  question("What is the danger of factors that look numeric (like 'var1', 'var10', 'var2')?",
    answer("They can't be plotted"),
    answer("They sort alphabetically ('var1', 'var10', 'var2') instead of numerically", correct = TRUE, message = "Correct! Alphanumeric sorting gives wrong order. Use gtools::mixedsort() to fix this."),
    answer("R will give an error"),
    answer("They automatically convert to numbers"),
    allow_retry = TRUE
  )
)
```

### Question 26

```{r q26}
quiz(
  question("What does forcats::fct_inorder() do?",
    answer("Sorts factors alphabetically"),
    answer("Sets factor levels in the order they first appear in data", correct = TRUE, message = "Correct! fct_inorder() preserves the order of first appearance, useful for pre-sorted data."),
    answer("Sorts factors numerically"),
    answer("Randomizes factor order"),
    allow_retry = TRUE
  )
)
```

### Question 27

```{r q27}
quiz(
  question("Which forcats function should you use to order factors based on another variable (like median)?",
    answer("fct_inorder()"),
    answer("fct_reorder()", correct = TRUE, message = "Correct! fct_reorder() orders factor levels based on a summary of another variable."),
    answer("fct_relevel()"),
    answer("fct_inseq()"),
    allow_retry = TRUE
  )
)
```

---

## Section 9: Interactive Plots

### Question 28

```{r q28}
quiz(
  question("What is the easiest way to convert a ggplot2 plot to interactive?",
    answer("Use shiny"),
    answer("ggplotly()", correct = TRUE, message = "Correct! ggplotly() from the plotly package instantly converts ggplot2 objects to interactive HTML widgets."),
    answer("plot.ly()"),
    answer("interactive_plot()"),
    allow_retry = TRUE
  )
)
```

### Question 29

```{r q29}
quiz(
  question("How do you save an interactive plot for sharing?",
    answer("ggsave()"),
    answer("htmlwidgets::saveWidget()", correct = TRUE, message = "Correct! saveWidget() saves interactive plots as self-contained HTML files."),
    answer("save()"),
    answer("You can't save interactive plots"),
    allow_retry = TRUE
  )
)
```

### Question 30

```{r q30}
quiz(
  question("What does selfcontained = TRUE do in saveWidget()?",
    answer("Deletes the plot after saving"),
    answer("Bundles all JavaScript/CSS into one HTML file", correct = TRUE, message = "Correct! selfcontained = TRUE creates a single file that can be easily shared via email."),
    answer("Compresses the HTML file"),
    answer("Requires internet to view"),
    allow_retry = TRUE
  )
)
```

### Question 31

```{r q31}
quiz(
  question("What is an advantage of ggiraph over ggplotly?",
    answer("It's faster"),
    answer("Better ggplot2 compatibility and theme preservation", correct = TRUE, message = "Correct! ggiraph is built specifically for ggplot2 and preserves themes better than ggplotly's conversion approach."),
    answer("It creates smaller files"),
    answer("It works with base R plots"),
    allow_retry = TRUE
  )
)
```

---

## Quiz Complete! 🎉

### How to Check Your Score

**Learnr tracks your progress automatically!**

Look at the **sidebar on the left** - you'll see:

- ✅ Green checkmarks for completed sections
- A progress indicator showing how many questions you've answered

To see which questions you got right or wrong:

- Scroll back through the quiz
- Questions with correct answers will show in **green**
- Questions with incorrect answers will show in **red** (if you attempted them)

### Score Interpretation Guide

**Estimate your performance based on the feedback you received:**

#### 🌟 Excellent (25-31 correct / 80%+)

You have a strong grasp of figure best practices! You're ready to apply these principles to your own work.

**Next steps:**

- Start implementing these practices in your projects
- Share this knowledge with your colleagues
- Bookmark the presentation as a reference

---

#### ✅ Good (19-24 correct / 60-79%)

You understand most concepts! A few areas need reinforcement.

**Next steps:**

- Review the sections where you missed questions
- Re-read the presentation slides for those topics
- Try the quiz again to reinforce learning

---

#### 📚 Needs Work (13-18 correct / 40-59%)

You're learning, but several concepts need more study.

**Next steps:**

- **Review the full presentation carefully**
- Focus on one topic at a time (colors, formats, etc.)
- **Retake the quiz** after reviewing
- Practice with your own data/plots

---

#### ⚠️ Please Retake (<13 correct / <40%)

These concepts are crucial for creating good scientific figures!

**What to do:**

1. **Take your time** - Don't rush through the presentation
2. **Review all slides** at [stanstrup.github.io/figure_presentation](https://stanstrup.github.io/figure_presentation/)
3. **Focus on understanding WHY** each principle matters (not just memorizing)
4. **Try applying concepts** to a real plot you've created
5. **Retake this quiz** once you've reviewed the material

**Important topics to master:**

- Why rainbow scales are problematic
- Vector vs raster formats
- When to use which file format
- How to save plots properly in R

### Key Takeaways

Remember the core principles:

1. **Colors**: Use viridis family, avoid rainbow
2. **Formats**: Vector (PDF/SVG) for plots, raster (PNG/TIFF) for photos
3. **Sizing**: Use smaller dimensions to make text larger
4. **Saving**: Use `ggsave()` with explicit dimensions and DPI
5. **Reproducibility**: Keep R scripts as source of truth
6. **Factors**: Watch out for alphanumeric sorting issues
7. **Interactive**: Use `ggplotly()` and save with `saveWidget()`

### What's Next?

**📚 Review the full presentation:**

[stanstrup.github.io/figure_presentation](https://stanstrup.github.io/figure_presentation/)

**🔄 Retake the quiz:**

Refresh this page or run `rmarkdown::run("quiz.Rmd")` again

**📖 Additional resources:**

- [ColorBrewer](https://colorbrewer2.org) - Color palette selection
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz) - Claus Wilke
- [The misuse of colour in science communication](https://www.nature.com/articles/s41467-020-19160-7)

---

### Feedback

Found an issue with the quiz? Have suggestions? Let us know!
