# Color Gradients {background-color="#e74c3c"}

## The Rainbow Problem

::: {.r-fit-text}
🌈 Why is rainbow/jet so popular if it's so bad?
:::

## Interactive Activity: Sort These Colors

**Task**: Arrange these colors from lowest to highest value

::: {.fragment}
🔴 Red | 🟡 Yellow | 🟢 Green | 🔵 Blue | 🟣 Purple | 🟠 Orange
:::

::: {.fragment}
**Can you do it intuitively?**

**NO!** And that's the problem.
:::

## The False Borders Problem

::: {.callout-warning}
## Real-World Example

Research shows rainbow colormaps create **false boundaries** in smooth data:

- Yellow-to-green transition "jumps out"
- Looks like sudden change where data is actually smooth
- Creates artificial "edges" in maps and heatmaps
:::

Source: *HESS, 2021* - "Rainbow color map distorts and misleads research in hydrology"

## Perceptual Non-Uniformity

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|5|10"

# Bad: Rainbow creates false artifacts
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_raster() +
  scale_fill_gradientn(colors = rainbow(7)) +
  labs(title = "Rainbow: Notice false 'edges'?")

# Good: Viridis is perceptually uniform
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_raster() +
  scale_fill_viridis_c() +
  labs(title = "Viridis: Smooth gradient")
```

## Medical Consequences

::: {.callout-important}
## From Research

"Physicians who use jet colormap in diagnosing heart disease take **longer** and make **significantly more errors** than those who use proper colormaps"

- Bright yellow appears more intense than dark red
- But dark red actually represents higher values!
:::

## Color Blindness

**Statistics:**
- ~8% of men
- ~0.5% of women
- Most common: Red-green deficiency

::: {.fragment}
**Classic failure**: Red-green heatmaps

These look identical to someone with deuteranopia!
:::

## Test Your Figures

::: {.incremental}
1. **Grayscale test**: Convert to B&W - can you still see patterns?
2. **Colorblind simulator**: [coblis.org](https://www.coblis.com) or ColorOracle
3. **Print test**: B&W printout still informative?
:::

## Recommended Palettes

::: {.columns}
::: {.column width="50%"}
**Sequential (low → high)**

```{r}
#| echo: true
#| eval: false
scale_color_viridis_c()
scale_color_brewer(
  palette = "Blues"
)
```
:::

::: {.column width="50%"}
**Diverging (- to +)**

```{r}
#| echo: true
#| eval: false
scale_color_brewer(
  palette = "RdBu"
)
scale_color_distiller(
  palette = "BrBG"
)
```
:::
:::

## Why Viridis Wins

::: {.incremental}
- ✅ Perceptually uniform (equal steps look equal)
- ✅ Colorblind-friendly
- ✅ Prints well in grayscale
- ✅ Increases monotonically in lightness
- ✅ No false boundaries or artifacts
:::

## Viridis Family

```{r}
#| echo: true
#| eval: false

# Four options in viridis family
scale_color_viridis_c(option = "viridis")  # blue-green-yellow (default)
scale_color_viridis_c(option = "magma")    # purple-red-yellow
scale_color_viridis_c(option = "plasma")   # purple-orange-yellow
scale_color_viridis_c(option = "inferno")  # black-purple-yellow
```

All are perceptually uniform!

## ColorBrewer

Great for categorical and diverging data

```{r}
#| echo: true
#| eval: false

# Categorical (up to 8-10 colors)
scale_color_brewer(palette = "Set1")
scale_color_brewer(palette = "Dark2")

# Sequential
scale_color_brewer(palette = "YlOrRd")

# Diverging
scale_color_brewer(palette = "RdYlBu")
```

[colorbrewer2.org](https://colorbrewer2.org)

## Demonstration: Rainbow vs. Viridis

```{r}
#| echo: true
#| eval: false

library(ggplot2)

# Create sample data with smooth gradient
df <- expand.grid(x = 1:100, y = 1:100)
df$z <- with(df, sin(x/10) * cos(y/10))

# Rainbow (bad)
p1 <- ggplot(df, aes(x, y, fill = z)) +
  geom_raster() +
  scale_fill_gradientn(colors = rainbow(7)) +
  labs(title = "Rainbow: See false borders?")

# Viridis (good)
p2 <- ggplot(df, aes(x, y, fill = z)) +
  geom_raster() +
  scale_fill_viridis_c() +
  labs(title = "Viridis: Smooth and perceptual")

library(patchwork)
p1 / p2
```

## Key Message

::: {.r-fit-text}
**Just Say No to Rainbow! 🚫🌈**
:::

Use viridis or ColorBrewer instead.
