# ggplot2 Themes {background-color="#9b59b6"}

## The Default Problem

```{r}
#| echo: true
#| eval: false

# Default ggplot2 theme
ggplot(mtcars, aes(wt, mpg)) +
  geom_point()  # Uses theme_gray()
```

::: {.fragment}
**Problems:**

- Gray background (wastes ink, unprofessional)
- Too much "chart junk"
- Not publication-ready
:::

## Built-in Clean Themes

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|4-5|7-8|10-11"

theme_bw()        # White background, black border
                  # Good for publications

theme_classic()   # No gridlines, clean axes
                  # Very minimal, classic journals

theme_minimal()   # Subtle gridlines, modern
                  # Good balance of clean + reference

theme_void()      # Blank canvas
                  # For custom designs, maps
```

## Visual Comparison

::: {.columns}
::: {.column width="50%"}
**theme_gray() (default)**

- Gray background
- White gridlines
- Bulky appearance
:::

::: {.column width="50%"}
**theme_classic()**

- White background
- No gridlines
- Clean, traditional
:::
:::

## Specialized Theme Packages

```{r}
#| echo: true
#| eval: false

# ggthemes package
library(ggthemes)
theme_few()         # Stephen Few's design principles
theme_tufte()       # Edward Tufte minimalist
theme_economist()   # The Economist magazine style
theme_fivethirtyeight()  # FiveThirtyEight website

# hrbrthemes package
library(hrbrthemes)
theme_ipsum()       # Modern professional typography
```

## Setting Global Theme

**Set once, apply to all plots:**

```{r}
#| echo: true
#| eval: false

# At top of script
theme_set(theme_bw(base_size = 12))

# Now all plots use theme_bw
ggplot(mtcars, aes(wt, mpg)) + geom_point()
ggplot(iris, aes(Sepal.Length, Sepal.Width)) + geom_line()
# Both use theme_bw automatically!
```

## Custom Theme for Publications

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1|3-4|5|6|7-8"

my_pub_theme <- theme_classic(base_size = 11) +
  theme(
    # Axis styling
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5),
    text = element_text(family = "Arial"),
    # Bold titles
    plot.title = element_text(face = "bold", size = 14)
  )

# Apply to plot
ggplot(mtcars, aes(wt, mpg)) + geom_point() + my_pub_theme
```

## Font Considerations

```{r}
#| echo: true
#| eval: false

# Check available fonts
library(extrafont)
font_import()  # First time only
fonts()        # List available

# Use in theme
theme_classic(base_family = "Arial") +
  theme(
    plot.title = element_text(family = "Arial", face = "bold")
  )
```

::: {.callout-note}
Many journals prefer Arial, Helvetica, or Times New Roman
:::

## Theme Elements You Can Customize

::: {.incremental}
- `axis.title` - Axis label text
- `axis.text` - Axis tick labels
- `legend.position` - "top", "bottom", "left", "right", "none"
- `panel.grid` - Gridlines
- `plot.background` - Entire plot area
- `strip.background` - Facet labels
:::

## Consistent Multi-Plot Style

```{r}
#| echo: true
#| eval: false

# Define once
pub_theme <- theme_classic(base_size = 11)
theme_set(pub_theme)

# Create multiple plots with same style
p1 <- ggplot(mtcars, aes(wt, mpg)) + geom_point()
p2 <- ggplot(iris, aes(Sepal.Length, Sepal.Width)) + geom_line()
p3 <- ggplot(faithful, aes(eruptions)) + geom_histogram()

# Combine into multi-panel figure
library(patchwork)
(p1 | p2) / p3
# All panels have consistent appearance!
```

## Quick Reference

| Theme | Use Case |
|-------|----------|
| `theme_classic()` | Traditional journals, minimal |
| `theme_bw()` | Modern publications, gridlines |
| `theme_minimal()` | Presentations, reports |
| `theme_void()` | Maps, custom designs |

## Recommendations

::: {.callout-tip}
## Best Practices

1. **Never use default** `theme_gray()` for publications
2. **Set global theme** at start of script for consistency
3. **Match journal style** - check published figures
4. **Keep it simple** - less chart junk = better
:::
