# Saving Plots in R {background-color="#34495e"}

```{r}
#| label: setup-saving
#| echo: false
#| eval: true
#| message: false
#| warning: false

library(ggplot2)
library(glue)
library(tidyverse)
library(ggpubr)
library(conflicted)
conflicts_prefer(dplyr::filter)

# Output directory
fig_dir <- "plots/06_saving"
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

## The Graphics Device System

**Every plot needs a "device"**

::: {.fragment}
Device = where R sends the graphics output (device = "the printer")

- Screen (RStudio viewer)
- PDF file
- SVG file
- PNG file
- JPEG file
- etc.
:::

## Old Way: Manual Device Management

```{r}
#| label: saving-old-way
#| echo: true
#| eval: false

pdf("myplot.pdf", width = 7, height = 5)
  plot(x, y)
dev.off()  # CRITICAL! Must close device
```

::: {.callout-note}
## dev.off() Works with ALL R Plots

The `dev.off()` approach works for:

- Base R plots (`plot()`, `hist()`, `barplot()`, etc.)
- ggplot2 plots
- Any R graphics output

**It's universal** - not limited to any specific plotting system!
:::

::: {.fragment}
**Problems with manual device management:**

- Easy to forget `dev.off()`
- Verbose
- Not intuitive
- Must manage device lifecycle manually
:::

## Modern Way: ggsave()

**For ggplot2 objects** (recommended!)

```{r}
#| label: saving-ggsave-basic
#| echo: true
#| eval: false

p <- ggplot(mtcars, aes(wt, mpg)) + geom_point()

# Saves last plot by default
ggsave("myplot.pdf")

# Better: explicit plot object
ggsave("myplot.pdf", plot = p, width = 6, height = 4)
```

No `dev.off()` needed! ✨

### Full Control

```{r}
#| label: saving-ggsave-full
#| echo: true
#| eval: false

ggsave(
  filename = "figure1.pdf",
  plot = my_plot,
  width = 7,
  height = 5,
  units = "in",     # or "cm", "mm"
  dpi = 300,        # for raster formats
  device = "pdf"    # or "png", "svg", "tiff"
)
```



## Batch Saving: Result

::: {.columns}
::: {.column width="50%"}

```{r}
#| label: saving-batch-show-data
#| echo: true
#| eval: false

# View the resulting tibble
plot_data
```

```
# A tibble: 3 × 3
  Species    data               plot
  <fct>      <list>             <list>
1 setosa     <tibble [50 × 4]>  <gg>
2 versicolor <tibble [50 × 4]>  <gg>
3 virginica  <tibble [50 × 4]>  <gg>
```

**Each row contains:**

- Species name
- Nested data for that species
- A ggplot object with regression

:::

::: {.column width="50%"}

```{r}
#| label: saving-batch-example-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5
#| message: false
#| warning: false

library(ggpubr)

# Example for setosa
iris_setosa <- iris %>% filter(Species == "setosa")

ggplot(iris_setosa, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 2, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred",
              formula = y ~ x) +
  stat_regline_equation(
    aes(label = after_stat(eq.label)),
    formula = y ~ x,
    label.x.npc = 0.98,
    label.y.npc = 0.95,
    hjust = 1
  ) +
  stat_cor(
    aes(label = paste(after_stat(rr.label), after_stat(p.label), sep = "~~~~")),
    label.x.npc = 0.98,
    label.y.npc = 0.88,
    hjust = 1
  ) +
  labs(title = "Iris setosa",
       x = "Sepal Length (cm)",
       y = "Sepal Width (cm)") +
  theme_classic(base_size = 11)
```

**Example:** Setosa species plot
:::
:::

## File Organization

```{r}
#| label: saving-organization
#| echo: true
#| eval: false

library(glue)

# Good practice: separate directory
fig_dir <- "figures"
dir.create(fig_dir, showWarnings = FALSE)

ggsave(glue("{fig_dir}/figure1.pdf"), p1, width = 7, height = 5)
ggsave(glue("{fig_dir}/figure1.png"), p1, width = 7, height = 5, dpi = 300)

# Save both vector and raster versions!
```



## Using magick for Conversion

```{r}
#| label: saving-magick-convert
#| echo: true
#| eval: false

library(magick)

# Convert PDF to 300 DPI PNG
img <- image_read_pdf("plot.pdf", density = 300)
image_write(img, "plot.png", format = "png", quality = 100)

# With better antialiasing (remove alpha channel)
img <- image_read_pdf("plot.pdf", density = 300)
img <- image_background(img, "white")  # Remove alpha
image_write(img, "plot.png", format = "png", quality = 100)

# Batch convert all PDFs in directory
pdf_files <- list.files(pattern = "\\.pdf$")
for (file in pdf_files) {
  img <- image_read_pdf(file, density = 300)
  img <- image_background(img, "white")
  out_file <- sub("\\.pdf$", ".png", file)
  image_write(img, out_file, format = "png", quality = 100)
}
```

## Batch Saving Multiple Plots

::: {.columns}
::: {.column width="60%"}


```{r}
#| label: saving-batch-fun
#| echo: true
#| eval: true

# Define plotting function
make_species_plot <- function(data, species) {
  ggplot(data, aes(x = Sepal.Length, y = Sepal.Width)) +
    geom_point(size = 2, color = "steelblue") +
    geom_smooth(method = "lm", se = TRUE, color = "darkred", formula = y ~ x) +
    
    stat_regline_equation(
      aes(label = after_stat(eq.label)),
      formula = y ~ x,
      label.x.npc = 0.98,  # 98% to the right (relative)
      label.y.npc = 0.95,  # 95% to the top (relative)
      hjust = 1            # right-align text
    ) +
    stat_cor(
      aes(label = paste(after_stat(rr.label), after_stat(p.label), sep = "~~~~")),
      label.x.npc = 0.98,
      label.y.npc = 0.88,
      hjust = 1
    ) +
    
    labs(title = glue("Iris {species}"), x = "Sepal Length (cm)", y = "Sepal Width (cm)") +
    theme_classic(base_size = 12)
}
```


```{r}
#| label: saving-batch
#| echo: true
#| eval: true

# Nest data by Species and apply plotting function
plot_data <- iris %>%
  nest(data = -Species) %>%
  mutate(plot = map2(data, Species, make_species_plot))
```

```{r}
#| label: saving-batch-save
#| echo: true
#| eval: false

# Save all plots using walk2
walk2(plot_data$plot, plot_data$Species, ~ggsave(
  filename = glue("{fig_dir}/iris_{..2}.pdf"),
  plot = ..1,
  width = 6, height = 5
))
```

:::

::: {.column width="40%"}

```{r}
#| label: saving-batch-tbl
#| echo: true
#| eval: true

plot_data
```

```{r}
#| label: saving-batch-show
#| echo: true
#| eval: true
#| fig-width: 5
#| fig-height: 3

plot_data$plot[[1]]
```

**Workflow:**

1. Nest iris data by Species
2. Define plotting function with regression
3. Use `map2()` to apply function
4. Save with `walk2()` using plot and Species columns to save one plot per species

:::
:::


## Recommendations

::: {.callout-tip}
## Best Practices

1. **Always use ggsave()** for ggplot2 (not manual devices)
2. **Always specify width, height, units and DPI (for raster output)**
3. **Increase DPI from the default** for all raster formats set ≥300
4. **Use cairo_pdf** for PDF output
5. **Save both PDF and PNG** versions
6. **Organize** figures in dedicated directory
:::
