# Factor Ordering {background-color="#2980b9"}

::: {.content-visible unless-format="revealjs"}
## Introduction

R's factor system is both powerful and treacherous. Alphabetical ordering makes categorical plots hard to read, and accidental conversion of numeric factors causes data corruption. This chapter teaches you how to control factor ordering and avoid common pitfalls.

### Learning Objectives

By the end of this chapter, you will:

- Understand why alphabetical factor ordering is rarely useful
- Learn to reorder factors by frequency, by another variable, or manually
- Master the `forcats` package for factor manipulation
- Know how to properly convert factors to numeric without data loss
- Avoid the "numeric factor" trap (numbered categories vs. actual numbers)
- Create publication-ready plots with meaningful orderings
:::

```{r}
#| label: setup-factors
#| echo: false
#| eval: true
#| message: false

library(ggplot2)
library(dplyr)
library(forcats)
library(patchwork)

# Set default theme for all plots
theme_set(theme_bw())
```

## The Problem: Alphabetical Ordering

::: {.columns}
::: {.column width="50%"}

```{r}
#| label: factors-problem-alphabetical
#| echo: true
#| eval: true

# Create data with logical categories
category_df <- data.frame(
  category = c("Low", "Medium", "High", "Very High", "Low", "High"),
  value = c(10, 20, 15, 30, 12, 25)
)

# R defaults to alphabetical!
p1 <- ggplot(category_df, aes(x = category, y = value)) +
  geom_col(fill = "coral") +
  theme_classic(base_size = 12) +
  labs(title = "Alphabetical (Wrong!)")
```

:::

::: {.column width="50%"}

```{r}
#| label: factors-solution-ordered
#| echo: true
#| eval: true

# Specify levels explicitly
category_df$category <- factor(category_df$category,
                       levels = c("Low", "Medium", "High", "Very High"))

# Now plots use logical order!
p2 <- ggplot(category_df, aes(x = category, y = value)) +
  geom_col(fill = "steelblue") +
  theme_classic(base_size = 12) +
  labs(title = "Logical Order (Correct!)")
```

:::
:::

::: {.columns}
::: {.column width="50%"}

```{r}
#| label: factors-problem-plot
#| echo: false
#| eval: true
#| fig-width: 5
#| fig-height: 3.5

p1
```

Random order makes no sense!

:::

::: {.column width="50%"}

```{r}
#| label: factors-solution-plot
#| echo: false
#| eval: true
#| fig-width: 5
#| fig-height: 3.5

p2
```

Much better - order makes sense!

:::
:::

## Solution: forcats Package

**Part of tidyverse, designed for factor manipulation**

::: {.columns}
::: {.column width="45%"}

```{r}
#| label: factors-forcats-intro
#| echo: true
#| eval: true

library(forcats)

# Reorder car classes by median highway mpg
p <- mpg %>%
  ggplot(aes(x = fct_reorder(class, hwy, median),
             y = hwy)) +
  geom_boxplot(fill = "lightblue") +
  coord_flip() +
  labs(x = "Vehicle Class",
       y = "Highway MPG",
       title = "Ordered by median MPG")
```

Boxplots now ordered by median value!

:::

::: {.column width="55%"}

```{r}
#| label: factors-forcats-intro-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

p
```

:::
:::

## fct_reorder(): Order by Another Variable

::: {.columns}
::: {.column width="45%"}

```{r}
#| label: factors-fct-reorder
#| echo: true
#| eval: true

# Order diamond cuts by mean price
p <- diamonds %>%
  group_by(cut) %>%
  summarise(mean_price = mean(price)) %>%
  ggplot(aes(x = fct_reorder(cut, mean_price),
             y = mean_price,
             fill = cut)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(x = "Diamond Cut",
       y = "Mean Price ($)",
       title = "Cuts ordered by average price")
```

Bars ordered by length!

:::

::: {.column width="55%"}

```{r}
#| label: factors-fct-reorder-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

p
```

:::
:::

## fct_infreq(): Order by Frequency

::: {.columns}
::: {.column width="45%"}

```{r}
#| label: factors-fct-infreq
#| echo: true
#| eval: true

# Order by how common each vehicle class is
p <- ggplot(mpg,
            aes(x = fct_infreq(class))) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  labs(x = "Vehicle Class",
       y = "Count",
       title = "Most common classes first")
```

Most common category first!

Great for survey data

:::

::: {.column width="55%"}

```{r}
#| label: factors-fct-infreq-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

p
```

:::
:::

## fct_inorder(): Order by Appearance

::: {.columns}
::: {.column width="45%"}

```{r}
#| label: factors-fct-inorder
#| echo: true
#| eval: true

# Create data with specific order
treatment_data <- data.frame(
  treatment = c("Control", "Low Dose",
                "Medium Dose", "High Dose",
                "Control", "Low Dose",
                "Medium Dose", "High Dose"),
  response = c(10, 12, 15, 18,
               11, 13, 16, 19)
)

# Keep order as they appear in data
p <- ggplot(treatment_data,
       aes(x = fct_inorder(treatment),
           y = response)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "Treatment", y = "Response")
```

Preserves the order from your data!

:::

::: {.column width="55%"}

```{r}
#| label: factors-fct-inorder-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

p
```

:::
:::

## Natural Sorting: var1, var2, ..., var10

**The problem with alphabetical sorting:**

::: {.columns}
::: {.column width="45%"}

```{r}
#| label: factors-natural-sort-problem
#| echo: true
#| eval: true

# Create data with numbered variables
var_data <- data.frame(
  variable = rep(c("var1", "var2", "var10", "var20"), each = 5),
  value = rnorm(20, mean = rep(c(10, 15, 20, 25), each = 5), sd = 2)
)

# Alphabetical order: var1, var10, var2, var20 (wrong!)
p <- ggplot(var_data, aes(x = variable, y = value)) +
  geom_boxplot(fill = "coral") +
  labs(title = "Alphabetical: var1, var10, var2, var20")
```

:::

::: {.column width="55%"}

```{r}
#| label: factors-natural-sort-problem-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

p
```

:::
:::

## Natural Sorting: The Solution

::: {.columns}
::: {.column width="45%"}

```{r}
#| label: factors-natural-sort-solution
#| echo: true
#| eval: true

# Use gtools::mixedsort() for natural/alphanumeric sorting
library(gtools)

var_data$variable <- factor(var_data$variable,
                            levels = mixedsort(unique(var_data$variable)))

# Natural order: var1, var2, var10, var20 (correct!)
p <- ggplot(var_data, aes(x = variable, y = value)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Natural sort: var1, var2, var10, var20")
```

**Note:** `forcats::fct_inseq()` only works if factor levels are purely numeric strings (e.g., "1", "2", "10"), not mixed alphanumeric like "var1", "var10"

:::

::: {.column width="55%"}

```{r}
#| label: factors-natural-sort-solution-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

p
```

:::
:::

## fct_rev(): Reverse Order

::: {.columns}
::: {.column width="45%"}

```{r}
#| label: factors-fct-rev
#| echo: true
#| eval: true

# Reverse the frequency order
p <- ggplot(mpg, aes(x = fct_rev(fct_infreq(class)))) +
  geom_bar(fill = "coral") +
  coord_flip() +
  labs(x = "Vehicle Class", y = "Count",
       title = "Least common classes first (reversed)")
```

Now least common first instead of most common!

:::

::: {.column width="55%"}

```{r}
#| label: factors-fct-rev-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

p
```

:::
:::

## fct_relevel(): Move Specific Levels

::: {.columns}
::: {.column width="45%"}

```{r}
#| label: factors-fct-relevel
#| echo: true
#| eval: true

# Move "Control" to front for treatment groups
treatment_data <- data.frame(
  treatment = c("Low Dose", "High Dose", "Control",
                "Medium Dose", "Low Dose", "Control"),
  response = c(12, 18, 10, 15, 13, 11)
)

p <- treatment_data %>%
  mutate(treatment = fct_relevel(treatment, "Control")) %>%
  ggplot(aes(treatment, response)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "Treatment", y = "Response")
```

Control always shown first

Common in experimental data!

:::

::: {.column width="55%"}

```{r}
#| label: factors-fct-relevel-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

p
```

:::
:::

## Facet Ordering

::: {.columns}
::: {.column width="45%"}

```{r}
#| label: factors-facet-ordering
#| echo: true
#| eval: true

# Order facets by median highway mpg for each vehicle class
p <- mpg %>%
  filter(class %in% c("pickup", "minivan", "compact")) %>%
  mutate(class = fct_reorder(class, hwy, median)) %>%
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  facet_wrap(~class) +
  labs(x = "Engine Displacement (L)", y = "Highway MPG")
```

Facet panels in meaningful order!

:::

::: {.column width="55%"}

```{r}
#| label: factors-facet-ordering-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

p
```

:::
:::

## ⚠️ The Danger of Numeric Factors

**Converting between factor and numeric can destroy your data!**

\

::: {.fragment}
::: {.columns}
::: {.column width="50%"}

```{r}
#| label: factors-numeric-danger-1
#| echo: true
#| eval: true

# Original numeric data
doses <- c(10, 20, 50, 100, 200)
print(doses)
```

:::

::: {.column width="50%"}

```{r}
#| label: factors-numeric-danger-1b
#| echo: true
#| eval: true

# Convert to factor (common in data import!)
dose_factor <- factor(doses)
print(dose_factor)
```

:::
:::
:::

\

::: {.fragment}
```{r}
#| label: factors-numeric-danger-1c
#| echo: true
#| eval: true

# Looks fine, right? But look at the internal structure:
str(dose_factor)
levels(dose_factor)
```
:::

\

::: {.fragment}
::: {.columns}
::: {.column width="50%"}

```{r}
#| label: factors-numeric-danger-2
#| echo: true
#| eval: true

# Try to convert back to numeric - WRONG!
as.numeric(dose_factor)
```

:::

::: {.column width="50%"}

```{r}
#| label: factors-numeric-danger-2b
#| echo: true
#| eval: true

# Correct way: convert via character
as.numeric(as.character(dose_factor))
```

:::
:::
:::

\

::: {.fragment}
**The danger:** Many functions silently convert factors to integers!
:::

## ⚠️ Missing Levels: The Silent Data Corruption

**Even worse: missing levels get renumbered!**



::: {.fragment}
::: {.columns}
::: {.column width="50%"}

```{r}
#| label: factors-missing-levels-1
#| echo: false
#| eval: true

# Subjects 1, 2, 4, 5 (subject 3 missing!)
subject <- c(1, 2, 4, 5, 1, 2, 4, 5)
response <- c(10, 15, 25, 30, 11, 16, 24, 29)

data.frame(subject = subject, response = response)
```

:::

::: {.column width="50%"}

```{r}
#| label: factors-missing-levels-2
#| echo: true
#| eval: true

# Convert to factor (happens during import!)
subject_factor <- factor(subject)
str(subject_factor)
```

Levels: "1" "2" "4" "5" - looks OK...

:::
:::
:::



::: {.fragment}
::: {.columns}
::: {.column width="50%"}

\

```{r}
#| label: factors-missing-levels-3
#| echo: true
#| eval: true

# Try to convert back - DATA CORRUPTION!
as.numeric(subject_factor)
```

**Returns:** 1 2 3 4 1 2 3 4

**Your subject 4 became 3!**

**Your subject 5 became 4!**

:::

::: {.column width="50%"}

\

::: {.fragment}
::: {.callout-warning}
## This is catastrophic for analysis!

Your statistical models and plots will use the wrong subject numbers. **Always use as.numeric(as.character(factor))** not as.numeric(factor). \
Or better yet. **NEVER** use numbers for categorical data!
:::
:::


:::
:::
:::



## ⚠️ Numeric Factors After Reordering

::: {.columns}
::: {.column width="50%"}

::: {.fragment}
```{r}
#| label: factors-numeric-danger-reordered-setup
#| echo: true
#| eval: true

dose_data <- tibble(
  dose = factor(c(0, 10, 50, 100, 200)),
  response = c(5, 25, 80, 70, 30)
) %>%
  mutate(dose_ordered =
         fct_reorder(dose, response))

dose_data
```

:::



::: {.fragment}
```{r}
#| label: factors-numeric-danger-reordered-convert
#| echo: true
#| eval: true

# Try to use numerically - DISASTER!
dose_data <- dose_data %>%
  mutate(
    wrong = as.numeric(dose_ordered),
    correct = as.numeric(as.character(dose_ordered))
  )
```

```{r}
#| label: factors-numeric-danger-reordered-convert-show
#| echo: false
#| eval: true

dose_data
```

:::

:::

::: {.column width="50%"}

::: {.fragment}
```{r}
#| label: factors-numeric-danger-reordered-plot
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 8

# Show the disaster
p1 <- ggplot(dose_data, aes(x = wrong, y = response)) +
  geom_point(size = 3, color = "red") +
  geom_line() +
  labs(title = "WRONG: as.numeric(factor)",
       x = "as.numeric(dose_ordered)",
       y = "Response") +
  theme(plot.title = element_text(color = "red", size = 11))

p2 <- ggplot(dose_data, aes(x = correct, y = response)) +
  geom_point(size = 3, color = "darkgreen") +
  geom_line() +
  labs(title = "CORRECT: as.numeric(as.character())",
       x = "Actual dose",
       y = "Response") +
  theme(plot.title = element_text(color = "darkgreen", size = 11))

p1 / p2
```
:::

:::
:::


## Recommendations: Avoid Numeric Factors

::: {.callout-warning}
## Best Practices

1. **Never store numeric values as factors** unless they represent categories
2. **Prefix categorical numbers**: Use "Group_1", "Group_2" instead of "1", "2"
3. **Check imported data**: CSV imports often convert numbers to factors
4. **Use `readr::read_csv()`** instead of `read.csv()` - better type detection and no implicit conversion to factors.
5. **Always convert via character**: `as.numeric(as.character(factor))` not `as.numeric(factor)`
6. **Check with `str()`** before analysis to verify data types
:::



::: {.fragment}
::: {.columns}
::: {.column width="50%"}

**Bad: Numeric categories**

```{r}
#| label: factors-numeric-bad-practice
#| echo: true
#| eval: true

# DON'T do this - ambiguous!
groups <- factor(c(1, 2, 4, 5))
str(groups)
# Are these numbers or categories?
```

:::

::: {.column width="50%"}

**Good: Prefixed categories**

```{r}
#| label: factors-numeric-good-practice
#| echo: true
#| eval: true

# DO this - clearly categorical!
groups <- factor(c("Group_1", "Group_2",
                   "Group_4", "Group_5"))
str(groups)
# Obviously categories, safe!
```

:::
:::
:::




## forcats Cheat Sheet

```{r}
#| label: factors-cheatsheet
#| echo: true
#| eval: false

library(forcats)

fct_reorder(f, x, fun)   # Order by another variable
fct_infreq(f)            # Order by frequency
fct_inorder(f)           # Order by appearance in data
fct_inseq(f)             # Order by numeric value (if purely numeric)
fct_rev(f)               # Reverse current order
fct_relevel(f, "A", "B") # Move specific levels to front
fct_recode(f, new = "old") # Rename levels
fct_lump_n(f, n = 5)     # Keep top n, lump others as "Other"
fct_explicit_na(f)       # Make NA a visible level

# For natural sort (var1, var2, var10):
factor(x, levels = gtools::mixedsort(unique(x)))
```


## Debugging Factor Issues

```{r}
#| label: factors-debugging
#| echo: true
#| eval: true

# Example: vehicle classes in mpg dataset
vehicle_class <- factor(mpg$class)

# Check level order
vehicle_class %>% levels()

# See factor structure (shows integer encoding)
str(vehicle_class)

# Check how many observations per level
table(vehicle_class)

# Convert to character if needed for text operations
class_char <- as.character(vehicle_class)
```

## Key Takeaways & Best Practices

::: {.callout-important}
## Remember

**The Problem:**

- **R defaults to alphabetical** factor ordering - usually wrong!
- **Numeric factors are dangerous** - converting destroys your data
- **Missing levels get renumbered** - group 4 becomes 3!

**The Solutions:**

- **forcats package** provides powerful ordering tools:
  - **`fct_reorder()`** orders by another variable (most useful!)
  - **`fct_infreq()`** orders by frequency
  - **`fct_relevel()`** to put control/baseline first
  - **`fct_rev()`** to reverse order
- **Manual levels** for logical ordering (Low/Med/High, months, etc.)
- **Prefix categorical numbers**: "Group_1" not "1"

**Always:**

- **Check factor order** before plotting with `str()` or `levels()`
- **Convert via character**: `as.numeric(as.character(f))` not `as.numeric(f)`
- **Think about your reader** - what order makes sense?
:::

::: {.content-visible unless-format="revealjs"}
## Summary

:::{.callout-tip}
## Key Takeaways

1. **Alphabetical order is rarely useful** - readers can't interpret patterns
2. **forcats package** provides essential tools:
   - `fct_reorder(category, value)` - order by another variable (most common!)
   - `fct_infreq()` - order by frequency (highest first)
   - `fct_rev()` - reverse current order
   - `fct_relevel(f, "Control")` - move specific level first
3. **Manual ordering** for logical sequences:
   ```r
   factor(x, levels = c("Low", "Medium", "High"))
   factor(month, levels = month.name)  # built-in constant
   ```
4. **The numeric factor trap**:
   - `as.numeric(factor(c("1", "2", "10")))` gives `1, 2, 3` (WRONG!)
   - `as.numeric(as.character(factor(c("1", "2", "10"))))` gives `1, 2, 10` (correct)
5. **Prefix numbered categories**: Use "Group_1" not "1" to prevent accidental conversion
6. **Always check**: `str(data)` and `levels(variable)` before plotting
7. **Think about your reader**: What ordering tells the story best?
:::

## Exercises

:::{.callout-note}
## Try It Yourself

1. Create a dataset with unordered categories and fix the ordering:
   ```r
   df <- data.frame(
     treatment = c("High", "Low", "Medium", "High", "Low"),
     response = c(8.2, 3.1, 5.5, 9.1, 2.8)
   )
   # Fix: factor(treatment, levels = c("Low", "Medium", "High"))
   ```
2. Practice `fct_reorder()`:
   ```r
   starwars %>%
     mutate(species = fct_reorder(species, height, .fun = median, na.rm = TRUE)) %>%
     ggplot(aes(height, species)) + geom_boxplot()
   ```
3. Create a factor from numbers ("1", "2", "10") and practice safe conversion
4. Compare plots with alphabetical vs. reordered factors - which is easier to read?
:::

## Further Reading

- [forcats package documentation](https://forcats.tidyverse.org/) - comprehensive guide
- [R for Data Science - Factors chapter](https://r4ds.hadley.nz/factors.html) - Hadley Wickham
- [Factor ordering in ggplot2](https://ggplot2.tidyverse.org/reference/scale_discrete.html) - limits parameter
- [The factor trap](https://www.r-bloggers.com/2020/02/factors-how-would-you-order-a-factor/) - common pitfalls
:::
