# Interactive Plots {background-color="#27ae60"}

## The Appeal of Interactivity

**Why use interactive plots?**

::: {.incremental}
- 🖱️ Hover to see exact values
- 🔍 Zoom and pan
- 👁️ Toggle traces on/off
- 📊 Great for exploration
- 🎯 Excellent for presentations
:::

## plotly and ggplotly

**Two approaches:**

::: {.columns}
::: {.column width="50%"}
**ggplotly()**

```{r}
#| echo: true
#| eval: false

library(plotly)

p <- ggplot(data, aes(x, y)) +
  geom_point()

ggplotly(p)
```

Convert ggplot2 to interactive!
:::

::: {.column width="50%"}
**plot_ly()**

```{r}
#| echo: true
#| eval: false

library(plotly)

plot_ly(data,
  x = ~x, y = ~y,
  type = 'scatter',
  mode = 'markers')
```

Build interactively from scratch
:::
:::

## Basic ggplotly Example

```{r}
#| echo: true
#| eval: false

library(ggplot2)
library(plotly)

# Create ggplot
p <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  labs(title = "Fuel Efficiency by Weight",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon",
       color = "Cylinders")

# Make interactive
ggplotly(p)
```

Try hovering over points!

## Customizing Tooltips

```{r}
#| echo: true
#| eval: false

# Control what appears on hover
p <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl),
                        text = paste("Car:", rownames(mtcars),
                                   "<br>HP:", hp))) +
  geom_point(size = 3)

ggplotly(p, tooltip = c("text", "x", "y"))

# Now hover shows car name and horsepower!
```

## The Saving Problem

::: {.callout-warning}
## Challenge

Interactive plots are **HTML widgets**, not static images!

Can't just save as PDF or PNG traditionally.
:::

## Solution 1: Save as HTML (Keep Interactivity)

```{r}
#| echo: true
#| eval: false

library(htmlwidgets)

# Create interactive plot
p_interactive <- ggplotly(my_plot)

# Save as self-contained HTML
saveWidget(p_interactive,
           "interactive_plot.html",
           selfcontained = TRUE)

# Can email this file or put on website!
# Opens in any web browser
```

## Solution 2: Static Snapshot

**Requires kaleido (Python)**

First, install kaleido:

```bash
# In terminal
pip install kaleido
```

Then in R:

```{r}
#| echo: true
#| eval: false

library(plotly)

# Save static image
save_image(p_interactive,
           "plot.png",
           width = 800,
           height = 600)

# Can also save as PDF, SVG, etc.
save_image(p_interactive, "plot.pdf")
```

## Solution 3: Screenshot (Quick & Dirty)

**In plotly viewer:**

1. Hover over plot
2. Click camera icon 📷 (top-right)
3. Downloads PNG

**Pros:** Quick

**Cons:** Not reproducible, fixed resolution

## When to Use Interactive vs. Static

::: {.columns}
::: {.column width="50%"}
**Use Interactive:**

- HTML reports (R Markdown)
- Quarto documents
- Lab presentations
- Exploratory analysis
- Supplementary materials
- Websites
:::

::: {.column width="50%"}
**Use Static:**

- Journal publications
- Print posters
- PowerPoint (sometimes)
- PDF reports
- When file size matters
:::
:::

## R Markdown Integration

```markdown
```{r, echo=FALSE, message=FALSE}
library(plotly)
p <- ggplot(data, aes(x, y)) + geom_point()
ggplotly(p)
```
```

When you knit to HTML, plot is interactive!

When you knit to PDF, automatically becomes static.

## Quarto Integration

```markdown
---
title: "My Report"
format: html
---

```{r}
#| label: interactive-plot
#| fig-cap: "Interactive scatter plot"

library(plotly)
ggplotly(my_plot)
```
```

Perfect for modern scientific reports!

## Advanced plotly Features

```{r}
#| echo: true
#| eval: false

# Add buttons for different views
p <- plot_ly(data, x = ~x, y = ~y, type = 'scatter', mode = 'markers')

p %>% layout(
  updatemenus = list(
    list(
      buttons = list(
        list(method = "restyle", args = list("type", "scatter"),
             label = "Scatter"),
        list(method = "restyle", args = list("type", "box"),
             label = "Box")
      )
    )
  )
)
```

## 3D Plots

```{r}
#| echo: true
#| eval: false

# 3D scatter plot
plot_ly(mtcars,
        x = ~wt, y = ~hp, z = ~mpg,
        color = ~factor(cyl),
        type = "scatter3d",
        mode = "markers")

# Rotate by clicking and dragging!
```

Great for exploring multivariate data

## Linked Plots (crosstalk)

```{r}
#| echo: true
#| eval: false

library(crosstalk)

# Create shared data
shared_data <- SharedData$new(mtcars)

# Two linked plots
p1 <- plot_ly(shared_data, x = ~wt, y = ~mpg, type = 'scatter')
p2 <- plot_ly(shared_data, x = ~hp, y = ~mpg, type = 'scatter')

# Display side-by-side
subplot(p1, p2)

# Selecting points in one highlights in both!
```

## Time Series with Range Slider

```{r}
#| echo: true
#| eval: false

# Stock price example
plot_ly(data, x = ~date, y = ~price, type = 'scatter', mode = 'lines') %>%
  layout(
    xaxis = list(
      rangeselector = list(
        buttons = list(
          list(count = 1, label = "1m", step = "month", stepmode = "backward"),
          list(count = 6, label = "6m", step = "month", stepmode = "backward"),
          list(count = 1, label = "1y", step = "year", stepmode = "backward"),
          list(step = "all", label = "All")
        )
      ),
      rangeslider = list(visible = TRUE)
    )
  )
```

## Animations

```{r}
#| echo: true
#| eval: false

# Animated scatter plot (e.g., by year)
plot_ly(gapminder,
        x = ~gdpPercap, y = ~lifeExp,
        size = ~pop, color = ~continent,
        frame = ~year,  # Animate over years
        text = ~country,
        type = 'scatter', mode = 'markers') %>%
  layout(xaxis = list(type = "log"))

# Play button appears automatically!
```

## File Size Considerations

**Interactive HTML can be large:**

```{r}
#| echo: true
#| eval: false

# For large datasets, sample for interactive version
data_sample <- data %>% slice_sample(n = 1000)

p <- ggplot(data_sample, aes(x, y)) + geom_point()
ggplotly(p)

# Keep full dataset for static publication figure
```

## Hybrid Approach: Both Versions

```{r}
#| echo: true
#| eval: false

# Create base plot
p_static <- ggplot(data, aes(x, y, color = group)) +
  geom_point(size = 3) +
  theme_classic(base_size = 12)

# Static version for publication
ggsave("figure1.pdf", p_static, width = 7, height = 5)

# Interactive version for supplement/website
p_interactive <- ggplotly(p_static)
saveWidget(p_interactive, "figure1_interactive.html")

# Best of both worlds!
```

## Sharing Interactive Plots

**Options:**

1. **Email HTML file** (if selfcontained = TRUE)
2. **Upload to web server**
3. **GitHub Pages** (free hosting)
4. **RPubs** (publish directly from RStudio)
5. **Shiny app** (for more complex interactions)

## Publishing to RPubs

```{r}
#| echo: true
#| eval: false

# In R Markdown document with interactive plots

# Click "Knit" button
# Then click "Publish" button
# Select RPubs
# Get public URL to share!
```

## Limitations of ggplotly

**Not all ggplot2 features convert:**

- Some geoms don't translate well
- Complex annotations may be lost
- Custom themes partially supported
- Facets work but can be slow

**Test your conversion!**

## Alternative: ggiraph

```{r}
#| echo: true
#| eval: false

library(ggiraph)

# Create plot with interactive elements
p <- ggplot(mtcars, aes(wt, mpg,
                        tooltip = rownames(mtcars),
                        data_id = rownames(mtcars))) +
  geom_point_interactive(size = 3)

girafe(ggobj = p)

# Hover to see tooltips
# Click to select
```

Better ggplot2 compatibility than plotly!

## Recommendations

::: {.callout-tip}
## Best Practices

1. **Interactive for exploration** and presentations
2. **Static for publications** and print
3. **Save HTML** to preserve interactivity
4. **Sample large datasets** to keep HTML file manageable
5. **Create both versions** when possible
6. **Test thoroughly** - not all features convert
7. **Consider audience** - do they have web browser?
:::

## Complete Example

```{r}
#| echo: true
#| eval: false

library(ggplot2)
library(plotly)
library(htmlwidgets)

# 1. Create publication-quality static plot
p_static <- ggplot(iris, aes(Sepal.Length, Sepal.Width,
                             color = Species)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 11) +
  labs(title = "Iris Measurements",
       x = "Sepal Length (cm)",
       y = "Sepal Width (cm)")

# 2. Save static version (for paper)
ggsave("iris_static.pdf", p_static, width = 5, height = 4)

# 3. Create interactive version (for supplement)
p_interactive <- ggplotly(p_static) %>%
  layout(hovermode = "closest")

# 4. Save interactive version
saveWidget(p_interactive, "iris_interactive.html", selfcontained = TRUE)

# Now you have both!
```

## Key Takeaways

::: {.callout-important}
## Remember

- Interactive plots are **HTML widgets**, not images
- Use `saveWidget()` to save with interactivity
- Use `save_image()` for static snapshot (requires kaleido)
- **Both formats have their place** - create both!
- Interactive excellent for **exploration**, static for **publication**
:::
