# Rendering Issues {background-color="#c0392b"}

## What You See Isn't Always What You Get

**Common scenario:**

::: {.incremental}
- Create plot in R
- Save as PDF
- Open in viewer
- **Looks broken!** 😱
- Lines missing, weird artifacts
:::

::: {.fragment}
But when you print or open in Illustrator... **it's perfect!**
:::

## Why This Happens

**Rendering vs. Content:**

::: {.columns}
::: {.column width="50%"}
**PDF contains:**

- Exact mathematical instructions
- Perfect vector data
- Precise coordinates
:::

::: {.column width="50%"}
**Viewer displays:**

- Approximation for screen
- Simplified for speed
- Varies by software
:::
:::

## The Thin Line Problem

```{r}
#| echo: true
#| eval: false

# Very thin lines may not render on screen
ggplot(mtcars, aes(wt, mpg)) +
  geom_line(linewidth = 0.1)  # Too thin!

# Looks broken in preview
# Prints fine
# Solution: use linewidth ≥ 0.5
```

## Transparency Rendering

```{r}
#| echo: true
#| eval: false

# Many semi-transparent points
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(alpha = 0.1, size = 1)

# 10,000 points at 10% transparency
# May appear as solid blob in some viewers
# May render slowly
# But prints correctly!
```

## Font Substitution

**Problem:**

::: {.incremental}
- You use specific font (e.g., "Roboto")
- Viewer doesn't have that font
- Substitutes different font
- Spacing/alignment looks wrong
:::

**Solution:**

```{r}
#| echo: true
#| eval: false

# Use cairo_pdf to embed fonts
ggsave("plot.pdf", p, width = 7, height = 5, device = cairo_pdf)
```

## Viewer Differences

**Same PDF, different appearance:**

| Viewer | Rendering Quality |
|--------|-------------------|
| Adobe Acrobat Reader | ⭐⭐⭐⭐⭐ Excellent |
| Preview (Mac) | ⭐⭐⭐⭐ Good |
| Chrome/Firefox | ⭐⭐⭐ Variable |
| Edge | ⭐⭐⭐ Variable |
| RStudio Viewer | ⭐⭐ Basic |

## Real-World Example

```{r}
#| echo: true
#| eval: false

# Create plot with many elements
p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_density2d(linewidth = 0.3) +
  theme_classic(base_size = 9) +
  theme(panel.border = element_rect(linewidth = 0.25, fill = NA))

ggsave("complex.pdf", p, width = 3.5, height = 3)

# In RStudio viewer: may look messy
# In Preview: looks OK
# In Acrobat: looks perfect
# When printed: looks perfect
```

## The Border Artifact Issue

**Some viewers show:**

- Jagged borders
- Gaps between shapes
- White lines between tiles

```{r}
#| echo: true
#| eval: false

# Heatmap with tiles
ggplot(mtcars, aes(wt, mpg, fill = hp)) +
  geom_tile()

# May show white lines between tiles in viewer
# These disappear when printed!
```

## Anti-Aliasing Differences

::: {.columns}
::: {.column width="50%"}
**Screen rendering:**

- Uses anti-aliasing
- Smooths edges
- Can create "fuzzy" appearance
- Varies by viewer settings
:::

::: {.column width="50%"}
**Print rendering:**

- Higher resolution (300+ DPI)
- Less/no anti-aliasing needed
- Crisp edges
- True to vector instructions
:::
:::

## Testing Your Figures

::: {.callout-tip}
## Best Practices

1. **Always check in Adobe Acrobat** (most accurate)
2. **Use print preview** (Command/Ctrl + P)
3. **Test actual printing** on paper
4. **Check journal's submission system** viewer
5. **When in doubt, trust print preview**
:::

## Avoiding Rendering Problems

```{r}
#| echo: true
#| eval: false

# 1. Use reasonable line widths
geom_line(linewidth = 0.5)  # Not 0.1

# 2. Use cairo_pdf for better compatibility
ggsave("plot.pdf", device = cairo_pdf)

# 3. Avoid excessive transparency layering
# Instead of 10,000 points at alpha=0.1:
geom_density_2d_filled()  # Computed density

# 4. Use standard fonts
theme_classic(base_family = "Arial")

# 5. Avoid too-small text
theme_classic(base_size = 10)  # Not 6
```

## Communication with Collaborators

**When sharing drafts:**

> "Please open in Adobe Acrobat Reader or check print preview. Some viewers may display rendering artifacts that aren't in the actual PDF."

## The "Looks Wrong" Checklist

**Before panicking:**

::: {.incremental}
1. ✅ Open in Adobe Acrobat Reader
2. ✅ Check print preview
3. ✅ View at 100% zoom (not zoomed way in)
4. ✅ Try printing one copy
5. ✅ Check on different device
:::

::: {.fragment}
**If it looks good in 3+ of these: it's fine!**
:::

## Specific Issues and Solutions

### Issue: Thin lines disappear

```{r}
#| echo: true
#| eval: false

# ❌ Bad
geom_line(linewidth = 0.1)

# ✅ Good
geom_line(linewidth = 0.5)  # Minimum for reliable rendering
```

### Issue: Fonts look wrong

```{r}
#| echo: true
#| eval: false

# ❌ Bad
ggsave("plot.pdf", device = "pdf")  # Doesn't embed fonts

# ✅ Good
ggsave("plot.pdf", device = cairo_pdf)  # Embeds fonts
```

## Advanced: Controlling PDF Version

```{r}
#| echo: true
#| eval: false

# Some journals require specific PDF version
cairo_pdf("plot.pdf",
          width = 7, height = 5,
          version = "1.4")  # PDF 1.4 (Acrobat 5.0)

print(my_plot)
dev.off()
```

Check journal requirements!

## Debugging Rendering Issues

```{r}
#| echo: true
#| eval: false

# Create simplified version to test
p_simple <- ggplot(head(mtcars, 20), aes(wt, mpg)) +
  geom_point(size = 3) +
  theme_classic(base_size = 12)

ggsave("test.pdf", p_simple, device = cairo_pdf)

# If simple version renders correctly:
# Problem is complexity (too many elements, transparency, etc.)

# If simple version still broken:
# Problem is likely viewer or device settings
```

## When to Worry (and When Not To)

::: {.columns}
::: {.column width="50%"}
**Don't Worry:**

- Thin lines look broken in RStudio
- Colors slightly different in browser
- Slight fuzziness on screen
- Minor spacing differences
:::

::: {.column width="50%"}
**Do Worry:**

- Missing data in Acrobat
- Can't read text when printed
- Actual graphical errors
- Elements overlapping incorrectly
:::
:::

## Summary

::: {.callout-important}
## Key Points

1. **Screen rendering ≠ print output**
2. **Different viewers = different rendering**
3. **Adobe Acrobat = most accurate** screen preview
4. **Print preview = ground truth**
5. **Use cairo_pdf** for best compatibility
6. **Avoid very thin lines** (< 0.5 pt)
7. **When in doubt, print it out!**
:::

## Trust the Print

::: {.r-fit-text}
If it prints well, it **is** well.
:::

::: {.fragment}
Don't obsess over screen artifacts!
:::
