# Factor Ordering {background-color="#2980b9"}

## The Alphabetical Problem

**R's default behavior:**

```{r}
#| echo: true
#| eval: false

categories <- c("High", "Medium", "Low", "Very High")
factor(categories)
# Levels: High Low Medium Very High

# Alphabetical! Not logical! 😞
```

## Visual Impact

::: {.columns}
::: {.column width="50%"}
**Alphabetical (bad):**

```
High
Low
Medium
Very High
```

Nonsensical!
:::

::: {.column width="50%"}
**Logical (good):**

```
Low
Medium
High
Very High
```

Makes sense!
:::
:::

## Why This Matters

::: {.incremental}
- **Bar charts**: Random bar order confuses readers
- **Legends**: Illogical color assignment
- **Line plots**: Colors don't match expectations
- **Facets**: Panels in wrong sequence
- **Boxplots**: Treatments in random order
:::

## Solution 1: Manual Ordering

```{r}
#| echo: true
#| eval: false

# Specify levels explicitly
data$category <- factor(data$category,
                       levels = c("Low", "Medium", "High", "Very High"))

# Now plots will use this order!
ggplot(data, aes(x = category, y = value)) +
  geom_col()
```

Full control over order

## Solution 2: forcats Package

**Part of tidyverse, designed for factor manipulation**

```{r}
#| echo: true
#| eval: false

library(forcats)

# Reorder by another variable (e.g., median)
data %>%
  mutate(group = fct_reorder(group, value, .fun = median)) %>%
  ggplot(aes(group, value)) +
  geom_boxplot()

# Boxplots now ordered by median value!
```

## fct_reorder(): Order by Another Variable

```{r}
#| echo: true
#| eval: false

# Order categories by mean of values
data %>%
  mutate(category = fct_reorder(category, value, mean)) %>%
  ggplot(aes(x = category, y = value, fill = category)) +
  geom_col() +
  coord_flip()  # Horizontal bars easier to read
```

Bars ordered by height!

## fct_infreq(): Order by Frequency

```{r}
#| echo: true
#| eval: false

# Order by how common each level is
ggplot(data, aes(x = fct_infreq(category))) +
  geom_bar() +
  labs(x = "Category") +
  coord_flip()

# Most common category first!
```

Great for survey data

## fct_rev(): Reverse Order

```{r}
#| echo: true
#| eval: false

# Sometimes you want reverse order
data %>%
  mutate(category = fct_reorder(category, value) %>% fct_rev()) %>%
  ggplot(aes(x = category, y = value)) +
  geom_col()

# Now highest values first instead of last
```

## fct_relevel(): Move Specific Levels

```{r}
#| echo: true
#| eval: false

# Move "Control" to front, keep others in order
data %>%
  mutate(treatment = fct_relevel(treatment, "Control")) %>%
  ggplot(aes(treatment, response)) +
  geom_boxplot()

# Control always shown first
```

Common in experimental data!

## Common Use Case: Ordered Bar Chart

```{r}
#| echo: true
#| eval: false

# Calculate summary statistics
summary_data <- data %>%
  group_by(name) %>%
  summarise(mean_value = mean(value))

# Plot with bars ordered by height
ggplot(summary_data,
       aes(x = fct_reorder(name, mean_value), y = mean_value)) +
  geom_col(fill = "steelblue") +
  coord_flip() +  # Horizontal bars
  labs(x = "", y = "Mean Value") +
  theme_minimal()
```

Much easier to read than alphabetical!

## Time-Based Ordering

```{r}
#| echo: true
#| eval: false

# Months in calendar order (not alphabetical!)
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

data$month <- factor(data$month, levels = months)

# Now plots show Jan, Feb, Mar... not alphabetical!
ggplot(data, aes(month, value)) +
  geom_line(group = 1)
```

## Legend Ordering

```{r}
#| echo: true
#| eval: false

# Order legend by maximum y value of each group
ggplot(data, aes(x = time,
                 y = value,
                 color = fct_reorder(group, value, max, .desc = TRUE))) +
  geom_line(linewidth = 1) +
  labs(color = "Group") +
  theme_minimal()

# Legend now shows groups in order of importance!
```

## Facet Ordering

```{r}
#| echo: true
#| eval: false

# Order facets by median value
data %>%
  mutate(category = fct_reorder(category, value, median)) %>%
  ggplot(aes(x = subcategory, y = value)) +
  geom_boxplot() +
  facet_wrap(~category) +
  theme_minimal()

# Facet panels in meaningful order!
```

## forcats Cheat Sheet

```{r}
#| echo: true
#| eval: false

library(forcats)

fct_reorder(f, x, fun)   # Order by another variable
fct_infreq(f)            # Order by frequency
fct_rev(f)               # Reverse current order
fct_relevel(f, "A", "B") # Move specific levels to front
fct_recode(f, new = "old") # Rename levels
fct_lump_n(f, n = 5)     # Keep top n, lump others as "Other"
fct_explicit_na(f)       # Make NA a visible level
```

## Real-World Example: Survey Data

```{r}
#| echo: true
#| eval: false

# Likert scale survey responses
survey <- data.frame(
  response = sample(c("Strongly Disagree", "Disagree", "Neutral",
                     "Agree", "Strongly Agree"), 100, replace = TRUE)
)

# Set logical order
survey$response <- factor(survey$response,
  levels = c("Strongly Disagree", "Disagree", "Neutral",
             "Agree", "Strongly Agree"))

# Plot with correct order
ggplot(survey, aes(x = response)) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  theme_minimal()
```

## Real-World Example: Treatment Groups

```{r}
#| echo: true
#| eval: false

library(forcats)

# Clinical trial data
clinical <- data.frame(
  treatment = rep(c("Placebo", "Low Dose", "Medium Dose", "High Dose"), each = 25),
  response = rnorm(100, mean = rep(c(10, 12, 15, 18), each = 25), sd = 3)
)

# Ensure logical ordering
clinical$treatment <- factor(clinical$treatment,
  levels = c("Placebo", "Low Dose", "Medium Dose", "High Dose"))

# Boxplots in dose order
ggplot(clinical, aes(treatment, response)) +
  geom_boxplot(fill = "lightblue") +
  theme_classic()
```

## Combining Operations

```{r}
#| echo: true
#| eval: false

# Combine multiple forcats functions
data %>%
  mutate(
    # 1. Make NA explicit
    category = fct_explicit_na(category),
    # 2. Lump rare categories
    category = fct_lump_n(category, n = 5),
    # 3. Order by frequency
    category = fct_infreq(category),
    # 4. Reverse so most common at top
    category = fct_rev(category)
  ) %>%
  ggplot(aes(category)) +
  geom_bar()
```

## Before and After

::: {.columns}
::: {.column width="50%"}
**Before (alphabetical):**

```{r}
#| echo: true
#| eval: false

ggplot(msleep,
  aes(x = vore,
      y = sleep_total)) +
  geom_boxplot()

# Order: carni, herbi,
#        insecti, omni
# Meaningless!
```
:::

::: {.column width="50%"}
**After (by median):**

```{r}
#| echo: true
#| eval: false

library(forcats)
ggplot(msleep,
  aes(x = fct_reorder(
           vore,
           sleep_total,
           median,
           na.rm = TRUE),
      y = sleep_total)) +
  geom_boxplot()

# Ordered by sleep time!
```
:::
:::

## Common Patterns

### Pattern 1: Sorted bar chart

```{r}
#| echo: true
#| eval: false

data %>%
  count(category) %>%
  ggplot(aes(x = fct_reorder(category, n), y = n)) +
  geom_col() +
  coord_flip()
```

### Pattern 2: Control group first

```{r}
#| echo: true
#| eval: false

data %>%
  mutate(group = fct_relevel(group, "Control")) %>%
  ggplot(aes(group, value)) +
  geom_boxplot()
```

### Pattern 3: Chronological facets

```{r}
#| echo: true
#| eval: false

data %>%
  mutate(year = fct_reorder(as.character(year), year_numeric)) %>%
  ggplot(aes(x, y)) +
  geom_point() +
  facet_wrap(~year)
```

## Debugging Factor Issues

```{r}
#| echo: true
#| eval: false

# Check current levels
levels(data$category)

# Check level order
data$category %>% levels()

# See factor structure
str(data$category)

# Convert to character if needed
data$category <- as.character(data$category)
```

## Recommendations

::: {.callout-tip}
## Best Practices

1. **Always check factor order** before plotting
2. **Use `fct_reorder()`** for data-driven ordering
3. **Use manual levels** for logical ordering (Low/Med/High)
4. **Use `fct_relevel()`** to put control/baseline first
5. **Use `fct_infreq()`** for categorical data
6. **Think about your reader** - what order makes sense?
:::

## Complete Example

```{r}
#| echo: true
#| eval: false

library(ggplot2)
library(dplyr)
library(forcats)

# Gene expression data by tissue type
expr_data <- data.frame(
  tissue = rep(c("Brain", "Heart", "Liver", "Lung", "Kidney"), each = 50),
  expression = c(rnorm(50, 100, 15), rnorm(50, 50, 10),
                 rnorm(50, 200, 30), rnorm(50, 75, 12),
                 rnorm(50, 150, 20))
)

# Plot with tissues ordered by median expression
expr_data %>%
  mutate(tissue = fct_reorder(tissue, expression, median)) %>%
  ggplot(aes(x = tissue, y = expression, fill = tissue)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic(base_size = 14) +
  labs(x = "Tissue Type",
       y = "Expression Level",
       title = "Gene Expression by Tissue")

# Tissues ordered by median expression level!
```

## Key Takeaways

::: {.callout-important}
## Remember

- **R defaults to alphabetical** factor ordering
- **forcats package** provides powerful ordering tools
- **`fct_reorder()`** orders by another variable
- **`fct_infreq()`** orders by frequency
- **Manual levels** for logical ordering (Low/Med/High)
- **Always think about the reader** - what order makes sense?
- **Check factor levels** before creating final plots
:::
