# Text and Element Sizing {background-color="#1abc9c"}

::: {.content-visible unless-format="revealjs"}
## Introduction

One of the most common mistakes in figure preparation is manually adjusting text sizes element by element until they "look right." This approach is inefficient, inconsistent, and defeats the purpose of using themes. This chapter reveals the elegant solution: control text size by adjusting figure dimensions, not font sizes.

### Learning Objectives

By the end of this chapter, you will:

- Understand how figure dimensions affect relative text size
- Know the standard figure widths for journals (single vs. double column)
- Learn to use `base_size` as the primary text sizing control
- Understand when and how to adjust individual element sizes
- Know the difference between DPI for vector vs. raster formats
- Be able to create publication-ready figures at the correct size on the first try
:::

```{r}
#| label: setup-sizing
#| echo: false
#| eval: true
#| message: false
#| warning: false

library(ggplot2)
library(patchwork)
library(glue)

# Output directory
fig_dir <- "plots/04_sizing"
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

# Create the base plot used throughout
p <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +

  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Weight (1000 lbs)", y = "Miles per Gallon",
       color = "Cylinders") +
  theme_classic()
```

## The Key Insight

::: {.callout-important}
## The Smart Way to Control Text Size

**Don't manually set font sizes for every element!**

Instead, use **smaller figure dimensions** to make text appear larger relative to the plot.

Then use **vector formats (SVG/PDF)** for infinite resolution.

:::

## Visual Demonstration

```{r}
#| label: sizing-demo-generate
#| echo: false
#| eval: true
#| output: false

# Save at different sizes to demonstrate the effect
ggsave(glue("{fig_dir}/sizing_small.svg"), p, width = 3.5, height = 3)
ggsave(glue("{fig_dir}/sizing_large.svg"), p, width = 7, height = 6)
```

::: {.columns}
::: {.column width="50%"}
**Small canvas (3.5 × 3 inches)**

![](plots/04_sizing/sizing_small.svg){width="100%"}

Text appears **large** relative to plot

```{r}
#| label: sizing-demo-generate2
#| echo: true
#| eval: false
#| code-overflow: wrap

ggsave("sizing_small.svg", p,
       width = 3.5, height = 3)
```

:::

::: {.column width="50%"}
**Large canvas (7 × 6 inches)**

![](plots/04_sizing/sizing_large.svg){width="100%"}

Text appears **small** relative to plot

```{r}
#| label: sizing-demo-generate3
#| echo: true
#| eval: false
#| code-overflow: wrap

ggsave("sizing_large.svg", p,
       width = 7, height = 6)
```

:::
:::


## base_size Effect (theme_classic(base_size = x))

```{r}
#| label: sizing-journal-demo
#| echo: false
#| eval: true

library(ggplot2)

# Create your plot with default theme
p <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Weight (1000 lbs)", y = "Miles per Gallon",
       color = "Cylinders") +
  theme_classic()

# STEP 1: Save at journal single-column width (3.5 inches)
# Text will appear appropriately large!
ggsave(glue("{fig_dir}/figure1.svg"), p, width = 3.5, height = 3)

# STEP 2: If text still too small/large, adjust base_size
p2 <- p + theme_classic(base_size = 10)  # Slightly smaller base
ggsave(glue("{fig_dir}/figure1_v2.svg"), p2, width = 3.5, height = 3)
```

```{r}
#| label: base-size-demo-generate
#| echo: false
#| eval: true
#| output: false

# Save with different base_size values
ggsave(glue("{fig_dir}/sizing_base8.svg"), p + theme_classic(base_size = 8), width = 3.5, height = 3)
ggsave(glue("{fig_dir}/sizing_base11.svg"), p + theme_classic(base_size = 11), width = 3.5, height = 3)
ggsave(glue("{fig_dir}/sizing_base14.svg"), p + theme_classic(base_size = 14), width = 3.5, height = 3)
```

::: {.columns}
::: {.column width="33%"}
**base_size = 8**

![](plots/04_sizing/sizing_base8.svg){width="100%"}
:::

::: {.column width="33%"}
**base_size = 11 (default)**

![](plots/04_sizing/sizing_base11.svg){width="100%"}
:::

::: {.column width="33%"}
**base_size = 14**

![](plots/04_sizing/sizing_base14.svg){width="100%"}
:::
:::

**Other elements scale relative to base_size:**

- `axis.title`: `1.1× base_size`
- `axis.text`: `0.8× base_size`
- `legend.text`: `0.8× base_size`

Use base_size as the PRIMARY adjustment - only customize individual elements if needed



## The Right Workflow

::: {.callout-tip}
## Three-Step Process

1. **Set dimensions to match final output size**
   - Journal single column: ~3.5 inches (Always check specific journal guidelines!)
   - Journal double column: ~7 inches
   - Presentation: ~10 inches

2. **Adjust `base_size` if needed**
   - Only fine-tune if text still too large/small

3. **Adjust individual elements if needed**
   - Use `theme()` to customize specific text sizes
   - Only if steps 1-2 don't achieve desired result
:::


### Why This Works

- Text/points/lines have **fixed** sizes
- If the canvas is small they appear larger



## When you must adjust some sizes individually


```{r}
#| label: sizing-manual-generate
#| echo: false
#| eval: true
#| output: false

# Without manual theme adjustments
p_default <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 2) +
  labs(x = "Weight (1000 lbs)", y = "MPG", color = "Cyl") +
  theme_classic(base_size = 11)

# With manual theme adjustments
p_custom <- p_default +
  theme(
        axis.title = element_text(size = 14, face = "bold"),
        axis.text =  element_text(size = 12),
        legend.title = element_text(size = 12, face = "bold"),
        legend.text =  element_text(size = 11)
  )

ggsave(glue("{fig_dir}/sizing_default_theme.svg"), p_default, width = 4, height = 3)
ggsave(glue("{fig_dir}/sizing_custom_theme.svg"), p_custom, width = 4, height = 3)
```

::: {.columns}
::: {.column width="50%"}
**Default (base_size = 11)**

![](plots/04_sizing/sizing_default_theme.svg){width="100%"}

```{r}
#| label: sizing-default-code
#| echo: true
#| eval: false
#| code-overflow: wrap

p + theme_classic(base_size = 11)
```
:::

::: {.column width="50%"}
**With manual adjustments**

![](plots/04_sizing/sizing_custom_theme.svg){width="100%"}

```{.r .code-clip}
p + theme_classic(base_size = 11) +
  theme(
    axis.title   = element_text(size = 14, face = "bold"),
    axis.text    = element_text(size = 12),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text  = element_text(size = 11)
  )
```
:::
:::




## Vector vs Raster: Key Differences

::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## Vector Formats (SVG, PDF)

**Dimensions control text/element proportions, not quality!**

- Scale infinitely without quality loss
- **DPI is ignored**
- What matters: aspect ratio & relative proportions
:::

```{.r style="font-size: 1.1em;"}
# Text larger in small.svg
ggsave("small.svg", p,
       width = 3.5, height = 3)
ggsave("large.svg", p,
       width = 7, height = 6)
```
:::

::: {.column width="50%"}
::: {.callout-warning}
## Raster Formats (PNG, TIFF)

**DPI controls pixel count and quality!**

- Fixed resolution, can pixelate when scaled
- **DPI matters** (300+ for print)
- pixels = width × DPI
:::

```{.r style="font-size: 1.1em;"}
# 1500×1200 pixels
ggsave("plot.png", p,
       width = 5, height = 4, dpi = 300)
# 360×288 pixels
ggsave("plot.png", p,
       width = 5, height = 4, dpi = 72)
```
:::
:::

::: {.content-visible unless-format="revealjs"}
## Summary

:::{.callout-tip}
## Key Takeaways

1. **Size figures to match final output** - not to match your screen
2. **Standard journal widths**:
   - Single column: ~3.5 inches (check specific journal!)
   - Double column: ~7 inches
   - Presentation: ~10 inches
3. **Use `base_size` as primary control** - all theme elements scale proportionally
4. **Smaller canvas = larger text** - text is fixed size, reducing plot area makes it relatively larger
5. **Vector formats ignore DPI** - SVG/PDF scale infinitely, dimensions control proportions
6. **Raster formats need high DPI** - PNG/TIFF at 300+ DPI for print quality
7. **Manual adjustments last** - only customize individual elements if dimensions + base_size aren't enough
8. **Test at final size** - view your saved figure at actual publication size to check readability
:::

## Exercises

:::{.callout-note}
## Try It Yourself

1. Create a plot and save it at three different widths:
   ```r
   p <- ggplot(mtcars, aes(mpg, hp)) + geom_point() + theme_classic()
   ggsave("single_col.pdf", p, width = 3.5, height = 3)
   ggsave("double_col.pdf", p, width = 7, height = 5)
   ggsave("slide.pdf", p, width = 10, height = 6)
   ```
2. Open each PDF and compare text sizes - notice how smaller canvas makes text appear larger
3. Try different `base_size` values (8, 11, 14, 18) at the same figure dimensions
4. Create a PNG at the same dimensions but different DPI values (72, 150, 300)
5. Look up your target journal's figure requirements - what are their single/double column widths?
:::

## Further Reading

- [ggplot2 theme reference](https://ggplot2.tidyverse.org/reference/theme.html) - all theme elements
- [Journal figure size requirements](https://www.elsevier.com/authors/policies-and-guidelines/artwork-and-media-instructions) - example from Elsevier
- [Understanding DPI vs resolution](https://www.prepressure.com/library/file-formats/jpeg) - technical details
- Wickham (2016). *ggplot2: Elegant Graphics for Data Analysis* - Chapter 8 on themes
:::
